The hedging hyperprior for the normal distribution

### 2023-03-12 text/23/normal-hedging-hyperprior.Rmd
Initial graphs for normal-hedging-hyperprior

$f(X) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$

Assume for simplicity that $\sigma$ is known. A more general case, with $\sigma$ unknown is described on a separate web page.

@@

$\tau = \frac{1}{\sigma}$

The parameter $\tau$ is called the precision. The normal density function in terms of $\tau$ is

$f(X) = \frac{\tau}{\sqrt{2 \pi}}e^{- \tau^2 (X-\mu)^2}$

Simplify this by removing any constants not involving $X$, $\mu$, or $\tau$.

$f(X) \propto \tau e^{- \tau^2 (X-\mu)^2}$

You have an informative prior on $\mu$ that also has a normal distribution

$g(\mu; \mu_0, \tau_0) \propto \tau_0 e^{-\tau_0^2 (\mu-\mu_0)^2}$

@@

$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2\sum(X_i-\mu)^2}$

@@

$L(\mathbf{X} | \mu) \propto \tau^n e^{-\sum \tau (X_i-\bar{X}+\bar{X}-\mu)^2}$

@@

$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2(\sum(X_i-\bar{X})^2} e^{-\tau^2 n (\bar{X}-\mu)^2}$

$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2((n-1)S^2} e^{-\tau^2 n (\bar{X}-\mu)^2}$

@@

$h(\mu | \mathbf{X}) \propto \tau_0 e^{-\tau_0^2(\mu-\mu_0)^2} \tau^n e^{-\tau((n-1)S^2} e^{-\tau^2\sum(\bar{X}-\mu)^2)}$

@@

$h(\mu | \mathbf{X}) \propto $

@@

$\frac{\tau n\bar{X}\beta + \tau_0 \mu_0}{n\tau+\tau0}$

can be written as a weighted average of the prior mean and the mean of the observed data.

The posterior precision is

$\tau0+n\tau$

```{r}
x_max <- 3

a0 <- 41
b0 <- 44

n <- 60
xbar <- 2.1

a1 <- a0 + n
b1 <- b0 + n*xbar

posterior_alpha <- a0+n
posterior_beta <- b0+n*xbar

@@

f <- function(x, theta) {
  theta^(-1) * 
    exp(-x/theta)
}

g <- function(theta, alpha, beta) {
  beta^alpha/gamma(alpha) *
    theta^(-alpha-1) * 
    exp(-beta/theta)
}

prior05 <- 1/qgamma(c(0.975), a0, b0) 
prior95 <- 1/qgamma(c(0.025), a0, b0)

post05 <- 1/qgamma(c(0.975), a1, b1) 
post95 <- 1/qgamma(c(0.025), a1, b1)

@@

Let's illustrate this with an informative prior with $\alpha$ = `r a0` and $\beta$ = `r b0`.

```{r}
g_grid <- g(x_grid, a0, b0)
outside <- x_grid < prior05 | x_grid > prior95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=g_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

```{r}
f_grid <- f(xbar, x_grid)^n
x_color <- "white"
data.frame(x=x_grid, y=f_grid) %>%
  ggplot(aes(x, y)) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

```{r}
h_grid <- g(x_grid, a1, b1)
outside <- x_grid < post05 | x_grid > post95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=h_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x_grid, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r sprintf("%.2f", b0/(a0-1))` and it excludes `r xbar`, the mean of the data.

@@

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

gstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)}
density <- outer(theta_grid, tau_grid, gstar)
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

It is hard to see at this scale, but the prior distribution is higher for small values of $\tau$ when $\theta4 is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r}
density <- outer(theta_grid, tau_grid, gstar)
density[11:40, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

fstar <- function(theta, tau) {f(xbar, theta)^n}
likelihood <- outer(theta_grid, tau_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

hstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)*f(xbar, theta)^n}
density <- outer(theta_grid, tau_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  theta_grid, 
  tau_grid, 
  density,
  labcex=1,
  levels=c(10, 25, 50, 75, 90)/100,
  xlab="theta",
  ylab="tau")
```

You can see that the hedging hyperprior did its job. The most probable values of $\tau$ are small and the prior mean (`r b0/(a0-1)`) is largely ignored in favor of the mean of the data (`r xbar`).

>>>>

$f(X; \mu, \sigma) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$

Assume for simplicity that $\sigma$ is known and equal to $\sigma_F$. A more general case, with $\sigma$ unknown is described on a separate web page.

@@

$\tau_F = \frac{1}{\sigma_F^2}$

The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance, The normal density function in terms of $\tau_F$ is

$f(X; \mu, \tau_F) = \sqrt{\frac{\tau_F}{2 \pi}}e^{- \frac{1}{2}\tau_F (X-\mu)^2}$

Simplify this by removing any constants not involving $X$, $\mu$, or $\tau_F$.

$f(X) \propto e^{- \frac{1}{2} \tau_F (X-\mu)^2}$

You have an informative prior on $\mu$ that also has a normal distribution with a prior mean $\mu_0$ and a prior precision $\tau_0$.

$g(\mu; \mu_0, \tau_0) \propto  e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2}$

For an informative prior distribution, $\tau_0$ is usually much smaller than $\tau_F$. The ratio $n_0=\tau_F/\tau_0$ is the effective prior sample size and is equivalent to adding $n_0$ pseudo data values of $\mu_0$ to the observed data.

@@

$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\mu)^2}$

@@

$L(\mathbf{X} | \mu) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\mu)^2}$

@@

$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\mu)^2}$

$L(\mathbf{X} | \mu) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\mu-\bar{X})^2}$

@@

$h(\mu | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\mu-\bar{X})^2)}$

[Note that the following formulas need a lot of work.]

Drop the middle exponential term, which is constant with respect to $\mu$. Set

$\mu_W=\frac{\tau_0 \mu_0 + n \tau_F \bar{X}}{\tau_0 + n \tau_F}$

and rewrite the other two exponentials as

$h(\mu | \mathbf{X}) \propto \tau_0 e^{-\tau_0^2(\mu - \mu_W + \mu_W - mu_0)^2} \tau_F^n e^{-\tau_F(n-1)S^2} e^{-n \tau_F^2(\mu - \mu_W + \mu_W - \bar{X})^2)}$

@@

$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)^2(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$

@@

$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0}$

and the posterior precision is

$n \tau_F + \tau_0$

The posterior mean can be written as a weighted average of the prior mean and the mean of the observed data.

```{r}
x_max <- 10

mu0 <- 8
tau0 <- 4

tauF <- 10

n <- 10
xbar <- 5

mu1 <- (tau0*mu0+n*tauF*xbar)/(tau0+n*tauF)
tau1 <- tau0+n*tauF

@@

f <- function(x, mu, tau) {
  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
}

g <- f

prior05 <- qnorm(0.025, mu0, tau0) 
prior95 <- qnorm(0.975, mu0, tau0)

post05 <- qnorm(0.025, mu1, tau1) 
post95 <- qnorm(0.025, mu1, tau1)

@@

Let's illustrate this with an informative prior with $\mu_0$ = `r mu0` and $\tau_0$ = `r tau0`.

```{r}
g_grid <- g(x_grid, mu0, tau0)
outside <- x_grid < prior05 | x_grid > prior95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=g_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

```{r}
f_grid <- f(xbar, x_grid, tauF)^n
x_color <- "white"
data.frame(x=x_grid, y=f_grid) %>%
  ggplot(aes(x, y)) + 
    geom_line() + xlab(" ") + ylab(" ")
```

@@

```{r}
h_grid <- g(x_grid, mu1, tau1)
outside <- x_grid < post05 | x_grid > post95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=h_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x_grid, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r sprintf("%.2f", mu0)` and it excludes `r xbar`, the mean of the data.

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

gstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF)+tauF)}
density <- outer(mu_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\mu$ is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r}
density <- outer(mu_grid, lambda_grid, gstar)
density[11:40, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(10, 25, 50, 75, 90)/100,
  xlab="mu",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r mu0`) is largely ignored in favor of the mean of the data (`r xbar`).
